Object-oriented implementations of the MPDATA advection equation solver in~C++,~Python~and~Fortran

### 2013-03-19 paper/doc/paper.tex
last changes before arXiv resubmission

@@

\input{formulae/preamble}

@@

@@

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of advection equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae/adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in an arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion.
  All relevant MPDATA formul\ae~are given in the text but are presented without
    derivation or detailed discussion.
  For a recent review of MPDATA-based techniques see \citet[][and references therein]{Smolarkiewicz_2006}.

@@

  All listings include line numbers printed to the left of the source code, with separate numbering for
    C++ (listings prefixed with C, black frame),
  \codecpp{code/cpp/listings.hpp}{listing00}{listing01}{1}
    Python (listings prefixed with P, blue frame) and
  \codepyt{code/pyt/listings.py}{listing00}{listing01}{1}
    Fortran (listings prefixed with F, red frame).
  \codefor{code/for/listings.f}{listing00}{listing01}{1}
  Programming language constructs when inlined in the text are 
    typeset in bold, e.g. \prog{GOTO 2}.

@@

  Double precision floating-point format is used in all three implementations.
  The codes begin with the following definitions:
  \codecpp{code/cpp/listings.hpp}{listing01}{listing02}{1}
  \codepyt{code/pyt/listings.py}{listing01}{listing02}{1}
  \codefor{code/for/listings.f}{listing01}{listing02}{1}
  which provide a convenient way of switching to different precision.

@@

  Presented C++ implementation of MPDATA is built upon the Blitz++ 
    library\footnote{Blitz++ is a C++ class library for scientific computing which uses the expression templates technique to achieve high performance, see \url{http://sf.net/projects/blitz/}}.
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects.
  Blitz++ is a header-only library\footnote{Blitz++ requires linking with \prog{libblitz} if debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code/cpp/listings.hpp}{listing02}{listing03}{1}
  Here \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} serve as alias identifiers 
    and are introduced in order to shorten the code.

  The power of Blitz++ comes from the ability to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ,,recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible.
  Luckily, the \prog{auto} return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code/cpp/listings.hpp}{listing03}{listing04}{1}
  The call to \prog{blitz::safeToReturn()} function is included in order to ensure that
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled, reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA the NumPy\footnote{NumPy is a Python package
    for scientific computing offering support for multi-dimensional arrays and a library
    of numerical algorithms, see \url{http://numpy.org/}} package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python code includes the following sequence of \prog{import} statements:
  \codepyt{code/pyt/listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, to match the settings of C++ and Fortran compilers used herein, the NumPy package is instructed 
    to ignore any floating-point errors, if such an option
    is available in the interpreter\footnote{\prog{numpy.seterr()} is not supported in PyPy as of version 1.9}.
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

@@

  In the C++ implementation the Boost\footnote{
    Boost is a free and open-source collection of peer-reviewed C++ libraries available at \url{http://www.boost.org/}.
    Several parts of Boost have been integrated into or inspired new additions to the C++ standard.
  } \prog{ptr\_vector} class is used to represent sequences of Blitz++ arrays 
    and at the same time to handle automatic freeing of dynamically allocated memory.
  The \prog{ptr\_vector} class is further customised by defining a derived structure which element-access \prog{[~]} 
    operator is overloaded with a modulo variant:
  \codecpp{code/cpp/listings.hpp}{listing04}{listing05}{1}
  Consequently the last element of any such sequence may be accessed at index \prog{-1}, the last but one at \prog{-2}, 
    and so on.

@@

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code/for/listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} type is defined solely for the purpose of overcoming the limitation 
    of lack of an array-of-arrays construct, and its only member field is a two-dimensional array.
  An array of \prog{arr\_t} is used hereinafter as a container for sequences of arrays.

@@

  In the C++ implementation first a global instance \prog{h} of an empty structure 
    \prog{hlf\_t} is defined, and then the plus and minus operators for \prog{hlf\_t} and \prog{rng\_t} are overloaded:
  \codecpp{code/cpp/listings.hpp}{listing05}{listing06}{1}
  This way, the arrays representing vector field components can be indexed using
    \prog{(i+h,j)}, \prog{(i-h,j)} etc. where \prog{h}~represents the half.

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code/pyt/listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code/pyt/listings.py}{listing04}{listing05}{1}

  In Fortran fractional array indexing is obtained through
    definition and instantiation of an object representing the half, and having appropriate
    operator overloads:
  \codefor{code/for/listings.f}{listing03}{listing04}{1}

@@

  An \prog{ext()} function is defined in all three implementation, in order to simplify 
    coding of array range extensions:
  \codecpp{code/cpp/listings.hpp}{listing06}{listing07}{1}
  \codepyt{code/pyt/listings.py}{listing05}{listing06}{1}
  \codefor{code/for/listings.f}{listing04}{listing05}{1}
  Consequently, a range depicted by $i\pm1/2$ may be expressed in the code as \prog{ext(i, h)}.
  In all three implementations the \prog{ext()} function accept the second
    argument to be an integer or a ''half'' (cf. section \ref{sec:grid}).

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of an order $d$ of a set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:
  \begin{equation*}\label{eq:pi}
    \input{formulae/pi}
  \end{equation*}

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}
    (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations):
  \codecpp{code/cpp/listings.hpp}{listing07}{listing08}{1}

  NumPy uses tuples of slices for addressing multi-dimensional array with a
    single object.
  Therefore, the following definition of function \prog{pi()} suffices to represent $\pi$:
  \codepyt{code/pyt/listings.py}{listing06}{listing07}{1}

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code/for/listings.f}{listing05}{listing06}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.11 and F.17--F.20).

@@

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=iii]
    \item{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item{allocating and deallocating the required memory,}
    \item{providing access to the solver state,}
    \item{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{enumerate}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code/cpp/listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with an updated (advected)
     values of \prog{psi[n]}.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

@@

  Python definition of the \prog{solver} class follows closely the C++ structure definition:
  \codepyt{code/pyt/listings.py}{listing07}{listing08}{1}
  The key difference stems from the fact that, unlike Blitz++, NumPy does not allow an array
    to have arbitrary index base -- in NumPy the first element is always addressed with 0.
  Consequently, while in C++ (and Fortran) the computational domain is chosen to start at (i=0, j=0)
    and hence a part of the halo region to have negative indices, in Python the halo region starts 
    at (0,0)\footnote{The reason to allow the domain to begin at an arbitrary index
    is mainly to ease debugging in case the code would be used in parallel computations
    using domain decomposition where each subdomain could have its own index base corresponding to
    the location within the computational domain}.
  However, since the whole halo logic is hidden within the solver, such details are not exposed to the
    user.
  The \prog{bcx} and \prog{bcy} boundary-condition specifications are passed to the solver through
    constructor-like \prog{\_\_init\_\_()} method as opposed to template parameters in C++.

  The above C++ and Python prototype solvers in principle allow to operate with any boundary condition
    objects that implement methods called from within the solver.
  This requirement is checked at compile-time in the case of C++, and at run-time in the case of Python.
  In order to obtain an analogous behaviour with Fortran, it is required to
    define, prior to definition of a solver type, 
    an abstract type with deferred procedures having abstract interfaces 
    \citep[sic!, see Table 2.1 in][for a summary of approximate correspondence of OOP 
      nomenclature between Fortran and C++]{Rouson_et_al_2012}:
  \codefor{code/for/listings.f}{listing06}{listing07}{1}
  Having defined the abstract type for boundary-condition objects, 
    a definition of a solver class following closely the C++ and Python counterparts may be provided:
  \codefor{code/for/listings.f}{listing07}{listing08}{1}

@@

  The solver definition described in section~\ref{sec:solver} requires a given
    boundary condition object to implement a \prog{fill\_halos()} method.
  An implementation of periodic boundary conditions in C++ is provided in the following listing:
  \codecpp{code/cpp/listings.hpp}{listing09}{listing10}{1}

@@

  MPDATA and donor-cell are explicit forward-in-time algorithms -- they allow to predict $\psi^{[n+1]}$ as a
    function of $\psi^{[n]}$ where $n$ and $n+1$ denote two adjacent time levels.
  The donor-cell scheme may be written as \citep[eq. 2 in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:donor}
    \input{formulae/donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:donor:F}
    \input{formulae/donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code/cpp/listings.hpp}{listing10}{listing11}{1}

  \noindent 
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code/cpp/listings.hpp}{listing11}{listing12}{1}

@@

  \codecpp{code/cpp/listings.hpp}{listing12}{listing13}{1}

@@

  As mentioned in the previous section, the donor-cell formula
    constitutes an advection scheme, hence we may use it to create
    a \prog{solver\_donorcell} implementation of the abstract \prog{solver} class: 
  \codecpp{code/cpp/listings.hpp}{listing13}{listing14}{1}
  The above definition is given as an example only.
  In the following sections an MPDATA solver of the same structure is defined.

@@

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step is a donor-cell step (eq.~\ref{eq:donor}) with the Courant number
    fields corresponding to the MPDATA antidiffusive velocities of the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:antidiff1}
    \input{formulae/antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:
  \begin{equation}\label{eq:antidiff2}
    \input{formulae/antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:
  \begin{equation}\label{eq:A}
    \input{formulae/A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae/B}
  \end{equation}

  If the denominator in equations \ref{eq:A} or \ref{eq:B} equals zero for a given {\em i} and {\em j}, 
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct (available in all three considered languages):
  \codecpp{code/cpp/listings.hpp}{listing14}{listing15}{1}

  The $A$ term defined in equation \ref{eq:A} takes the following form:
  \codecpp{code/cpp/listings.hpp}{listing15}{listing16}{1}

  The $B$ term defined in equation \ref{eq:B} takes the following form:
  \codecpp{code/cpp/listings.hpp}{listing16}{listing17}{1}

  Equation \ref{eq:antidiff2} takes the following form:
  \codecpp{code/cpp/listings.hpp}{listing17}{listing18}{1}

  Equation \ref{eq:antidiff1} take the following form:
  \codecpp{code/cpp/listings.hpp}{listing18}{listing19}{1}

@@

  An MPDATA solver may be now constructed by inheriting from \prog{solver} class
    with the following definition in C++:
  \codecpp{code/cpp/listings.hpp}{listing19}{listing20}{1}
  The array of sequences of temporary arrays \prog{tmp} allocated in the constructor
    is used to store the antidiffusive velocities from the present and optionally 
    previous timestep (if using more than two iterations).

@@

  \codecpp{code/cpp/plot.cpp}{listing20}{listing21}{1}

>>>>

%%%%%\documentclass[review,peprint,10pt,5p,times]{elsarticle}
%%%%%\usepackage[switch]{lineno}
%%%%%\addtolength{\hoffset}{.5cm}

@@

\input{formulae_preamble}

@@

%%%%%  \linenumbers % for the review

%%%%%  \section*{Program summary}

%%%%%  \noindent
%%%%%  Program title: mpdata-oop\\
%%%%%  Licensing provisions: GNU GPL version 3 \citep{GPLv3}\\
%%%%%  Programming language: C++11, Python and Fortran 2008\\
%%%%%  External routines/libraries used: Blitz++, NumPy\\
%%%%%  Nature of problem: advective transport\\
%%%%%  Solution method: MPDATA\\
%%%%%  References: Smolarkiewicz 1984: A fully multidimensional positive definite advection transport algorithm with small implicit diffusion. J. Comp. Phys. 54, 325--362 \citep{Smolarkiewicz_1984}

@@

  MPDATA \citep{Smolarkiewicz_1984} stands for Multidimensional Positive Definite Advective Transport Algorithm and is
    an example of a numerical procedure used in weather, climate and ocean simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002}.
  MPDATA is a solver for systems of advection equations of the following form:
  \begin{equation}\label{eq:adv}
    \input{formulae_adv}
  \end{equation}
  that describe evolution of a scalar field $\psi$ transported by the fluid flow with velocity $\vec{v}$.
  Quoting Numerical Recipes once more, development of methods to numerically solve such problems 
    {\em ''is an art as much as a science''} \citep[][Sec.~20.1]{Press_et_al_2007},
    and MPDATA is an example of the state-of-the art in this field.
  MPDATA is designed to accurately solve equation (\ref{eq:adv}) in an arbitrary
    number of dimensions assuring positive-definiteness of scalar field $\psi$ 
    and incurring small numerical diffusion.
  All relevant MPDATA formul\ae~are given in the text but are presented without
    derivation or detailed discussion.
  For a recent review of MPDATA-based techniques see \citet[][and references therein]{Smolarkiewicz_2006}.

@@

  All listings include line numbers printed to the left of the source code, with separate numbering for
    C++ (listings prefixed with C, black frame),
  \codecpp{code-cpp-listings.hpp}{listing00}{listing01}{1}
    Python (listings prefixed with P, blue frame) and
  \codepyt{code-pyt-listings.py}{listing00}{listing01}{1}
    Fortran (listings prefixed with F, red frame).
  \codefor{code-for-listings.f}{listing00}{listing01}{1}
  Programming language constructs when inlined in the text are 
    typeset in bold, e.g. \prog{GOTO 2}.

@@

  Double precision floating-point format is used in all three implementations.
  The codes begin with the following definitions:
  \codecpp{code-cpp-listings.hpp}{listing01}{listing02}{1}
  \codepyt{code-pyt-listings.py}{listing01}{listing02}{1}
  \codefor{code-for-listings.f}{listing01}{listing02}{1}
  which provide a convenient way of switching to different precision.

@@

  Presented C++ implementation of MPDATA is built upon the Blitz++ 
    library\footnote{Blitz++ is a C++ class library for scientific computing which uses the expression templates technique to achieve high performance, see \url{http://sf.net/projects/blitz/}}.
  Blitz offers object-oriented representation of n-dimensional arrays,
    and array-valued mathematical expressions.
  In particular, it offers loop-free notation for array arithmetics
    that does not incur creation of intermediate temporary objects.
  Blitz++ is a header-only library\footnote{Blitz++ requires linking with \prog{libblitz} if debugging mode is used} -- to use it, it is enough to include the appropriate header file,
    and optionally expose the required classes to the present namespace:
  \codecpp{code-cpp-listings.hpp}{listing02}{listing03}{1}
  Here \prog{arr\_t}, \prog{rng\_t} and \prog{idx\_t} serve as alias identifiers 
    and are introduced in order to shorten the code.

  The power of Blitz++ comes from the ability to express array expressions as objects.
  In particular, it is possible to define a function that returns an array expression;
    i.e. not the resultant array, but an object representing a ,,recipe'' defining the operations
    to be performed on the arguments.
  As a consequence, the return types of such functions become unintelligible.
  Luckily, the \prog{auto} return type declaration from the C++11 standard allows to simplify the code significantly,
    even more if used through the following preprocessor macro:
  \codecpp{code-cpp-listings.hpp}{listing03}{listing04}{1}
  The call to \prog{blitz::safeToReturn()} function is included in order to ensure that
    all arrays involved in the expression being returned continue to exist in the
    caller scope.
  For example, definition of a function returning its array-valued argument doubled, reads:
    \prog{auto~f(arr\_t~x)~return\_macro(2*x)}.
  This is the only preprocessor macro defined herein.

  For the Python implementation of MPDATA the NumPy\footnote{NumPy is a Python package
    for scientific computing offering support for multi-dimensional arrays and a library
    of numerical algorithms, see \url{http://numpy.org/}} package is used.
  In order to make the code compatible with both the standard CPython
    as well as the alternative PyPy implementation of Python \citep[][]{Bolz_et_al_2011},
    the Python code includes the following sequence of \prog{import} statements:
  \codepyt{code-pyt-listings.py}{listing02}{listing03}{1}
  First, the PyPy's built-in NumPy implementation named \prog{numpypy} is imported if applicable (i.e. if running PyPy), 
    and the lazy evaluation mode is turned on through the \prog{set\_invalidation(False)} call.
  PyPy's lazy evaluation obtained with the help of a just-in-time compiler enables to achieve
    an analogous to Blitz++ temporary-array-free handling of array-valued expressions 
    (see discussion in section~\ref{sec:perf}).
  Second, to match the settings of C++ and Fortran compilers used herein, the NumPy package is instructed 
    to ignore any floating-point errors, if such an option
    is available in the interpreter\footnote{\prog{numpy.seterr()} is not supported in PyPy as of version 1.9}.
  The above lines conclude all code modifications that needed to be added in order to run
    the code with PyPy.

@@

  In the C++ implementation the Boost\footnote{
    Boost is a free and open-source collection of peer-reviewed C++ libraries available at \url{http://www.boost.org/}.
    Several parts of Boost have been integrated into or inspired new additions to the C++ standard.
  } \prog{ptr\_vector} class is used to represent sequences of Blitz++ arrays 
    and at the same time to handle automatic freeing of dynamically allocated memory.
  The \prog{ptr\_vector} class is further customised by defining a derived structure which element-access \prog{[~]} 
    operator is overloaded with a modulo variant:
  \codecpp{code-cpp-listings.hpp}{listing04}{listing05}{1}
  Consequently the last element of any such sequence may be accessed at index \prog{-1}, the last but one at \prog{-2}, 
    and so on.

@@

  Fortran does not feature any built-in sequence container capable of storing arrays, 
    hence a custom \prog{arrvec\_t} type is introduced:
  \codefor{code-for-listings.f}{listing02}{listing03}{1}
  The \prog{arr\_t} type is defined solely for the purpose of overcoming the limitation 
    of lack of an array-of-arrays construct, and its only member field is a two-dimensional array.
  An array of \prog{arr\_t} is used hereinafter as a container for sequences of arrays.

@@

  In the C++ implementation first a global instance \prog{h} of an empty structure 
    \prog{hlf\_t} is defined, and then the plus and minus operators for \prog{hlf\_t} and \prog{rng\_t} are overloaded:
  \codecpp{code-cpp-listings.hpp}{listing05}{listing06}{1}
  This way, the arrays representing vector field components can be indexed using
    \prog{(i+h,j)}, \prog{(i-h,j)} etc. where \prog{h}~represents the half.

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code-pyt-listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code-pyt-listings.py}{listing04}{listing05}{1}

  In Fortran fractional array indexing is obtained through
    definition and instantiation of an object representing the half, and having appropriate
    operator overloads:
  \codefor{code-for-listings.f}{listing03}{listing04}{1}

@@

  An \prog{ext()} function is defined in all three implementation, in order to simplify 
    coding of array range extensions:
  \codecpp{code-cpp-listings.hpp}{listing06}{listing07}{1}
  \codepyt{code-pyt-listings.py}{listing05}{listing06}{1}
  \codefor{code-for-listings.f}{listing04}{listing05}{1}
  Consequently, a range depicted by $i\pm1/2$ may be expressed in the code as \prog{ext(i, h)}.
  In all three implementations the \prog{ext()} function accept the second
    argument to be an integer or a ''half'' (cf. section \ref{sec:grid}).

  \subsection{Array index permutations}\label{sec:pi}
  Hereinafter, the $\pi_{a,b}^{d}$ symbol is used to denote a  
    cyclic permutation of an order $d$ of a set $\{a,b\}$.
  It is used to generalise the MPDATA formul\ae~into multiple dimensions
    using the following notation:
  \begin{equation*}\label{eq:pi}
    \input{formulae_pi}
  \end{equation*}

  Blitz++ ships with the \prog{RectDomain} class (aliased here as \prog{idx\_t}) 
    for specifying array ranges in multiple dimensions.
  The $\pi$ permutation is implemented in C++ as a function \prog{pi()}
    returning an instance of \prog{idx\_t}.
  In order to ensure compile-time evaluation, the permutation order is passed 
    via the template parameter \prog{d}
    (note the different order of \prog{i} and \prog{j} arguments in the two template specialisations):
  \codecpp{code-cpp-listings.hpp}{listing07}{listing08}{1}

  NumPy uses tuples of slices for addressing multi-dimensional array with a
    single object.
  Therefore, the following definition of function \prog{pi()} suffices to represent $\pi$:
  \codepyt{code-pyt-listings.py}{listing06}{listing07}{1}

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code-for-listings.f}{listing05}{listing06}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.11 and F.17--F.20).

@@

  The tasks to be handled by a prototype advection equation solver proposed herein are:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=iii]
    \item{storing arrays representing the $\psi$ and $\vec{C}$ fields and any required housekeeping data,}
    \item{allocating and deallocating the required memory,}
    \item{providing access to the solver state,}
    \item{performing the integration by invoking the advection-operator and
      boundary-condition handling routines.}
  \end{enumerate}
  In the following C++ definition of the \prog{solver} structure, task (i) is represented with the definition of the structure
    member fields; task (ii) is split between the \prog{solver}'s constructor and the destructors of \prog{arrvec\_t};
    task (iii) is handled by the accessor methods; task (iv) is handled within the \prog{solve} method:
  \codecpp{code-cpp-listings.hpp}{listing08}{listing09}{1}
  The \prog{solver} structure is an abstract definition (containing a pure virtual method) 
     requiring its descendants to implement at least
     the \prog{advop()} method which is expected to fill \prog{psi[n+1]} with an updated (advected)
     values of \prog{psi[n]}.
  The two template parameters \prog{bcx\_t} and \prog{bcy\_t} allow the solver to operate with 
    any kind of boundary condition structures that fulfil the requirements implied by the calls to the 
    methods of \prog{bcx} and \prog{bcy}, respectively.

@@

  Python definition of the \prog{solver} class follows closely the C++ structure definition:
  \codepyt{code-pyt-listings.py}{listing07}{listing08}{1}
  The key difference stems from the fact that, unlike Blitz++, NumPy does not allow an array
    to have arbitrary index base -- in NumPy the first element is always addressed with 0.
  Consequently, while in C++ (and Fortran) the computational domain is chosen to start at (i=0, j=0)
    and hence a part of the halo region to have negative indices, in Python the halo region starts 
    at (0,0)\footnote{The reason to allow the domain to begin at an arbitrary index
    is mainly to ease debugging in case the code would be used in parallel computations
    using domain decomposition where each subdomain could have its own index base corresponding to
    the location within the computational domain}.
  However, since the whole halo logic is hidden within the solver, such details are not exposed to the
    user.
  The \prog{bcx} and \prog{bcy} boundary-condition specifications are passed to the solver through
    constructor-like \prog{\_\_init\_\_()} method as opposed to template parameters in C++.

  The above C++ and Python prototype solvers in principle allow to operate with any boundary condition
    objects that implement methods called from within the solver.
  This requirement is checked at compile-time in the case of C++, and at run-time in the case of Python.
  In order to obtain an analogous behaviour with Fortran, it is required to
    define, prior to definition of a solver type, 
    an abstract type with deferred procedures having abstract interfaces 
    \citep[sic!, see Table 2.1 in][for a summary of approximate correspondence of OOP 
      nomenclature between Fortran and C++]{Rouson_et_al_2012}:
  \codefor{code-for-listings.f}{listing06}{listing07}{1}
  Having defined the abstract type for boundary-condition objects, 
    a definition of a solver class following closely the C++ and Python counterparts may be provided:
  \codefor{code-for-listings.f}{listing07}{listing08}{1}

@@

  The solver definition described in section~\ref{sec:solver} requires a given
    boundary condition object to implement a \prog{fill\_halos()} method.
  An implementation of periodic boundary conditions in C++ is provided in the following listing:
  \codecpp{code-cpp-listings.hpp}{listing09}{listing10}{1}

@@

  MPDATA and donor-cell are explicit forward-in-time algorithms -- they allow to predict $\psi^{[n+1]}$ as a
    function of $\psi^{[n]}$ where $n$ and $n+1$ denote two adjacent time levels.
  The donor-cell scheme may be written as \citep[eq. 2 in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:donor}
    \input{formulae_donor}
  \end{equation}
  where $N$ is the number of dimensions, 
    and F is the so-called flux function \citep[][eq.~3]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:donor:F}
    \input{formulae_donor-F}
  \end{equation}

  \noindent The flux function takes the following form in C++:
  \codecpp{code-cpp-listings.hpp}{listing10}{listing11}{1}

  \noindent 
  Equation~\ref{eq:donor} is split into the terms under the summation 
    (effectively the 1-dimensional donor-cell formula):
  \codecpp{code-cpp-listings.hpp}{listing11}{listing12}{1}

@@

  \codecpp{code-cpp-listings.hpp}{listing12}{listing13}{1}

@@

  As mentioned in the previous section, the donor-cell formula
    constitutes an advection scheme, hence we may use it to create
    a \prog{solver\_donorcell} implementation of the abstract \prog{solver} class: 
  \codecpp{code-cpp-listings.hpp}{listing13}{listing14}{1}
  The above definition is given as an example only.
  In the following sections an MPDATA solver of the same structure is defined.

@@

  MPDATA introduces corrective steps to the algorithm defined by equation~\ref{eq:donor} and \ref{eq:donor:F}.
  Each corrective step is a donor-cell step (eq.~\ref{eq:donor}) with the Courant number
    fields corresponding to the MPDATA antidiffusive velocities of the following 
    form \citep[eqs~13,~14~in][]{Smolarkiewicz_1984}:
  \begin{equation}\label{eq:antidiff1}
    \input{formulae_antidiff1}
  \end{equation}
  where $\psi$ and $C$ represent values from the previous iteration and where:
  \begin{equation}\label{eq:antidiff2}
    \input{formulae_antidiff2}
  \end{equation}
  For positive-definite $\psi$, the $A$ and $B$ terms take the 
  following form\footnote{
    Since $\psi\ge0$, $|A|\le1$ and $|B|\le1$. 
    See \citet[Sec. 4.2]{Smolarkiewicz_2006} for description of adaptation of the 
    formul\ae~for advection of fields of variable sign
  }:
  \begin{equation}\label{eq:A}
    \input{formulae_A}
  \end{equation}
  \begin{equation}\label{eq:B}
    \input{formulae_B}
  \end{equation}

  If the denominator in equations \ref{eq:A} or \ref{eq:B} equals zero for a given {\em i} and {\em j}, 
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct (available in all three considered languages):
  \codecpp{code-cpp-listings.hpp}{listing14}{listing15}{1}

  The $A$ term defined in equation \ref{eq:A} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing15}{listing16}{1}

  The $B$ term defined in equation \ref{eq:B} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing16}{listing17}{1}

  Equation \ref{eq:antidiff2} takes the following form:
  \codecpp{code-cpp-listings.hpp}{listing17}{listing18}{1}

  Equation \ref{eq:antidiff1} take the following form:
  \codecpp{code-cpp-listings.hpp}{listing18}{listing19}{1}

@@

  An MPDATA solver may be now constructed by inheriting from \prog{solver} class
    with the following definition in C++:
  \codecpp{code-cpp-listings.hpp}{listing19}{listing20}{1}
  The array of sequences of temporary arrays \prog{tmp} allocated in the constructor
    is used to store the antidiffusive velocities from the present and optionally 
    previous timestep (if using more than two iterations).

@@

  \codecpp{code-cpp-plot.cpp}{listing20}{listing21}{1}
